Research log

### 2025-10-09 docs/ExploratoryResults.Rmd
model fits merged into one file, saving intermediate data automatically, and beautifying some parts of the paper

classoption       : "man"
output            : word_document

@@

```{r global_options, include=FALSE}
set.seed(42)
knitr::opts_chunk$set(fig.crop = F, fig.pos = "tb", fig.path='figs/',
                      echo=F, warning=F, cache=T, 
                      message=F, sanitize = T, 
                      fig.width = 10, fig.height = 5)
```

@@

@@

groundhog.library(c(
  'png',
  'grid',
  'ggplot2',
  'svglite',
  'xtable',
  'papaja',
  'tidyverse',
  'broom',
  'cowplot',
  'MESS', # for AUCs
  'lsr', # for effect sizes
  'pwr', # for power calculations
  'brms', # for mixed effects modeling
  'BayesFactor', # for Bayesian t test
  'jsonlite', # parsing data from sort_trial
  'caret', #for cross validation
  'ggrepel' #for word scatterplots
), '2025-06-01')

@@

```{r hangman-load-coefficients, echo=FALSE}

#get list of subjects
E4.data_for_turing <- read.csv('../analysis/dataForTuring/E4norm.csv')
E4.subjects <- E4.data_for_turing$subj_id%>%unique()

#initialize
E4.model_coefficients_raw <- data.frame()

#fill the array
for (i_subj in 1:length(E4.subjects)) {
  E4.model_coefficients_raw <- E4.model_coefficients_raw %>%
    rbind(read.csv(sprintf('../analysis/model_fits/E4/subj%d_normalized.csv',i_subj)) %>%
            mutate(subj_id=E4.subjects[i_subj]) %>%
            dplyr::select(subj_id,term,estimate))
}

# Organize in a more convenient form
# non-pretend dataframe

E4.np_coefficients <- E4.model_coefficients_raw %>% 
  group_by(subj_id) %>% 
  summarize(test_part = 'nonpretend',
            prior = estimate[term == 'coef_prior'],
            post = estimate[term == 'coef_post'],
            eig = estimate[term == 'coef_eig'])

E4.p_coefficients <- E4.model_coefficients_raw %>% 
  group_by(subj_id) %>% 
  summarize(test_part='pretend',
            prior = estimate[term == 'coef_prior'] +
              estimate[term == 'coef_prior_delta'],
            post = estimate[term == 'coef_post'] +
              estimate[term == 'coef_post_delta'],
            eig = estimate[term == 'coef_eig'] +
              estimate[term == 'coef_eig_delta'])

E4.model_coefficients <- E4.np_coefficients %>%
  rbind(E4.p_coefficients)
```

@@

E4.model_coefficients %>%
  gather('term','coefficient',prior,post,eig) %>%
  mutate(test_part = factor(test_part, levels=c('nonpretend','pretend'),
                            labels=c('non-pretend','pretend')),
         term = factor(term, levels = c('prior','post','eig'), labels = c('prior','posterior','expected information gain'))) %>%
  ggplot(aes(x=coefficient,fill=test_part, color=test_part))+
  geom_histogram( alpha=0.6, position = 'identity', bins=100) +
  scale_fill_manual(values=c("#69b3a2", "#404080")) +
  scale_color_manual(values=c("#146464", "#404080")) +
  labs(x='coefficient (a.u.)', y='number of players')+
  theme(legend.position=c(0.10,0.85)) +
  scale_x_continuous(limits=c(-2,5)) +
  facet_wrap(~term, nrow=3)+
  # scale_x_reverse() +
  theme_classic();

@@

All three terms had a positive effect in both conditions. Prior probability (base rate) had a stronger effect in pretend games (`r apa_print(E4.model_coefficients %>% dplyr::select(subj_id,test_part,prior)%>%spread(test_part,prior)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`). Hit posterior probability showed no such difference, and had a similar effect on players' decisions in both conditions (`r apa_print(E4.model_coefficients %>% dplyr::select(subj_id,test_part,post)%>%spread(test_part,post)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`). Finally, expected information gain had a weaker effect in pretend games (`r apa_print(E4.model_coefficients %>% dplyr::select(subj_id,test_part,eig)%>%spread(test_part,eig)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`).

```{r hangman-load-coefficients-orth, echo=FALSE}

#get list of subjects

#initialize
E4.model_coefficients_orth_raw <- data.frame()

#fill the array
for (i_subj in 1:length(E4.subjects)) {
  E4.model_coefficients_orth_raw <- E4.model_coefficients_orth_raw %>%
    rbind(read.csv(sprintf('../analysis/model_fits/E4/subj%d_orth.csv',i_subj)) %>%
            mutate(subj_id=E4.subjects[i_subj]) %>%
            dplyr::select(subj_id,term,estimate))
}

# Organize in a more convenient form
# non-pretend dataframe

E4.np_orth_coefficients <- E4.model_coefficients_orth_raw %>% 
  group_by(subj_id) %>% 
  summarize(test_part = 'nonpretend',
            prior = estimate[term == 'coef_prior'],
            post = estimate[term == 'coef_post'],
            eig = estimate[term == 'coef_eig'])

E4.p_orth_coefficients <- E4.model_coefficients_orth_raw %>% 
  group_by(subj_id) %>% 
  summarize(test_part='pretend',
            prior = estimate[term == 'coef_prior'] +
              estimate[term == 'coef_prior_delta'],
            post = estimate[term == 'coef_post'] +
              estimate[term == 'coef_post_delta'],
            eig = estimate[term == 'coef_eig'] +
              estimate[term == 'coef_eig_delta'])

E4.model_coefficients_orth <- E4.np_orth_coefficients %>%
  rbind(E4.p_orth_coefficients)
```

Given collinearity between model predictors, we suspected that the effects of expected information gain may be driven by posterior probability effects. To isolate unique contributions of prior probability and expected information gain, we repeated the same analysis, but this time serially orthogonalizing the features (in the order [posterior, prior, expected information gain], and using the Gram Schmidt algorithm) prior to fitting the multinomial regression model. This time, expected information gain had no significant effect on decisions in non-pretend (`r  apa_print(E4.model_coefficients_orth%>%filter(test_part=='nonpretend')%>%pull(eig)%>%t.test())$statistic`) and pretend games (`r  apa_print(E4.model_coefficients_orth%>%filter(test_part=='pretend')%>%pull(eig)%>%t.test())$statistic`), rendering it possible that the previously observed effects were confounded with prior and posterior probability. In contrast, both prior and posterior probability had a significant positive effect, with prior probability having a stronger effect in pretend games (`r apa_print(E4.model_coefficients_orth %>% dplyr::select(subj_id,test_part,prior)%>%spread(test_part,prior)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`).

@@

E4.model_coefficients_orth %>%
  gather('term','coefficient',prior,post,eig) %>%
  mutate(test_part = factor(test_part, levels=c('nonpretend','pretend'),
                            labels=c('non-pretend','pretend')),
         term = factor(term, levels = c('prior','post','eig'), labels = c('prior','posterior','expected information gain'))) %>%
  ggplot(aes(x=coefficient,fill=test_part, color=test_part))+
  geom_histogram( alpha=0.6, position = 'identity', bins=100) +
  scale_fill_manual(values=c("#69b3a2", "#404080")) +
  scale_color_manual(values=c("#146464", "#404080")) +
  labs(x='coefficient (a.u.)', y='number of players', title='Orthogonalized vectors')+
  theme(legend.position=c(0.10,0.85)) +
  scale_x_continuous(limits=c(-2,5)) +
  facet_wrap(~term, nrow=3)+
  # scale_x_reverse() +
  theme_classic();

@@

\newpage

>>>>

classoption       : "man"
output            : bookdown::gitbook

@@

```{r global_options, include=FALSE}
set.seed(42)
knitr::opts_chunk$set(fig.crop = F, fig.pos = "tb", fig.path='figs/',
                      echo=F, warning=F, cache=T, 
                      message=F, sanitize = T, 
                      fig.width = 10, fig.height = 5,
                      dev = "ragg_png", dpi = 300, fig.width = 6, fig.height = 4)
```

@@

# Run on R 4.2.2 ("Innocent and Trusting")

@@

groundhog.library(c(
  'png',
  'grid',
  'ggplot2',
  'svglite',
  'xtable',
  'papaja',
  'tidyverse',
  'broom',
  'cowplot',
  'MESS', # for AUCs
  'lsr', # for effect sizes
  'pwr', # for power calculations
  'brms', # for mixed effects modeling
  'BayesFactor', # for Bayesian t test
  'jsonlite', # parsing data from sort_trial
  'caret', #for cross validation
  'ggrepel', #for word scatterplots
  'ragg' # for png
), '2022-12-01')

@@

```{r hangman-load-coefficients, echo=FALSE}
E4.model_coefficients <- read.csv('../analysis/model_fits/model_coefficients.csv')
```

@@

E4.model_coefficients %>%
  filter(type=='basic') %>%
  gather('term','coefficient',prior,post,eig) %>%
  mutate(test_part = factor(test_part, levels=c('nonpretend','pretend'),
                            labels=c('non-pretend','pretend')),
         term = factor(term, levels = c('prior','post','eig'), labels = c('prior','posterior','expected information gain'))) %>%
  ggplot(aes(x=coefficient,fill=test_part, color=test_part))+
  geom_histogram( alpha=0.6, position = 'identity', bins=100) +
  scale_fill_manual(values=c("#69b3a2", "#404080")) +
  scale_color_manual(values=c("#146464", "#404080")) +
  labs(x='coefficient (a.u.)', y='number of players')+
  theme(legend.position=c(0.10,0.85)) +
  scale_x_continuous(limits=c(-2,5)) +
  facet_wrap(~term, nrow=3)+
  # scale_x_reverse() +
  theme_classic();

@@

All three terms had a positive effect in both conditions. Prior probability (base rate) had a stronger effect in pretend games (`r apa_print(E4.model_coefficients %>% filter(type=='basic') %>% dplyr::select(subj_id,test_part,prior)%>%spread(test_part,prior)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`). Hit posterior probability showed no such difference, and had a similar effect on players' decisions in both conditions (`r apa_print(E4.model_coefficients %>% filter(type=='basic') %>% dplyr::select(subj_id,test_part,post)%>%spread(test_part,post)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`). Finally, expected information gain had a weaker effect in pretend games (`r apa_print(E4.model_coefficients %>% filter(type=='basic') %>% dplyr::select(subj_id,test_part,eig)%>%spread(test_part,eig)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`).

Given collinearity between model predictors, we suspected that the effects of expected information gain may be driven by posterior probability effects. To isolate unique contributions of prior probability and expected information gain, we repeated the same analysis, but this time serially orthogonalizing the features (in the order [posterior, prior, expected information gain], and using the Gram Schmidt algorithm) prior to fitting the multinomial regression model. This time, expected information gain had no significant effect on decisions in non-pretend (`r  apa_print(E4.model_coefficients %>% filter(type=='orth') %>%filter(test_part=='nonpretend')%>%pull(eig)%>%t.test())$statistic`) and pretend games (`r  apa_print(E4.model_coefficients%>% filter(type=='orth') %>%filter(test_part=='pretend')%>%pull(eig)%>%t.test())$statistic`), rendering it possible that the previously observed effects were confounded with prior and posterior probability. In contrast, both prior and posterior probability had a significant positive effect, with prior probability having a stronger effect in pretend games (`r apa_print(E4.model_coefficients %>% filter(type=='orth') %>% dplyr::select(subj_id,test_part,prior)%>%spread(test_part,prior)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`).

@@

E4.model_coefficients %>%
  filter(type=='orth') %>%
  gather('term','coefficient',prior,post,eig) %>%
  mutate(test_part = factor(test_part, levels=c('nonpretend','pretend'),
                            labels=c('non-pretend','pretend')),
         term = factor(term, levels = c('prior','post','eig'), labels = c('prior','posterior','expected information gain'))) %>%
  ggplot(aes(x=coefficient,fill=test_part, color=test_part))+
  geom_histogram( alpha=0.6, position = 'identity', bins=100) +
  scale_fill_manual(values=c("#69b3a2", "#404080")) +
  scale_color_manual(values=c("#146464", "#404080")) +
  labs(x='coefficient (a.u.)', y='number of players', title='Orthogonalized vectors')+
  theme(legend.position=c(0.10,0.85)) +
  scale_x_continuous(limits=c(-2,5)) +
  facet_wrap(~term, nrow=3)+
  # scale_x_reverse() +
  theme_classic();

@@

```{r save data, echo-FALSE}

save.image(file = file.path(getwd(), "explor_results.RData"))

```
\newpage
